---
title: "Car Transmission Impact on Fuel Economy"
author: "Martin Tapp"
date: "February 27, 2016"
output: html_document
---

### Summary
This report summarizes the impact of car transmission type on fuel economy. It found that manual transmissions yield better fuel economy when compared to automatic transmissions with an average of 2.94 more miles per gallon. The results were derived from the Motor Trend Car Road Tests (`mtcars`). The full data processing is provided for anyone wishing to reproduce these results.


### Exploratory Data Analysis

Looking at the `mtcars` data, we can see that there are 11 features and 32 data points. We can see that the data contains a number of features which can be treated as factors to simplify our analysis (notably `cyl`, `vs`, `am`, `gear` and `carb`).

```{r, results='hide'}
data(mtcars)
dim(mtcars) # results hidden
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
```

#### Basic Transmission Type Influence

```{r}
fit_am <- lm(mpg ~ am, mtcars)
round(summary(fit_am)$adj.r.squared * 100, 1)
round(fit_am$coefficients, 1)
```

Let's look at the influence of transmission type (`am`) on `mpg` using a very basic model. We see that the automatic transmission (`intercept`, i.e., `am0`) gives an average of 17.1 mpg while the manual transmission (`am1`) yields an additinal 7.2 mpg over automatic transmissions. Unfortunately, this model only explains 33.8% of the `mpg` variance based on the adjusted R-squared. Therefore, we need to build a better model in order to assert the influence of transmission type on fuel economy.

#### Better Model

```{r}
fit_all <- lm(mpg ~ ., mtcars)
round(summary(fit_all)$adj.r.squared * 100, 1)
```

Let's fit a linear model consisting of all the features used to predict `mpg`. We can see that this model explains 77.9% of the `mpg` variance. After inspecting the summary, we notice that this model exhibits high standard errors for some features. Therefore, let's use the `step` function to derive a better model which contains the most influencial features for predicting `mpg`.

```{r, results='hide'}
fit_best <- step(fit_all, k = log(nrow(mtcars))) # results hidden
```
```{r}
summary(fit_best)$call
round(summary(fit_best)$adj.r.squared * 100, 1)
```

This yields a model involving weight, quarter mile time, and transmission type for predicting `mpg`. Our new model explains 83.4% of the `mpg` variance. Comparing our three models clearly shows that `fit_best` is the best one as it is under it's significant level (Pr(F) of 0.91 < 1).

```{r, results='hide'}
anova(fit_am, fit_all, fit_best) # results hidden
```


### Insights

```{r}
round(fit_best$coefficients, 2)
```
```{r, results='hide'}
round(confint(fit_best), 2) # results hidden
```

From our model, a manual transmission yields better mileage than an automatic transmission. This is supported by looking at the intercept of the best model. We see that manual transmissions yield 2.94 more miles per gallon than automatics when keeping weight and quarter mile time constant. We also see this increase is between 0.05 and 5.83 mpg 95% of the time.


### Appendix

#### Transmission Type influence

The following figure presents the influence of transmission type (`am`) on fuel economy (`mpg`) using a simplified linear model. We clearly see the benefit of manual over automatic transmissions.

```{r, results='hide'}
boxplot(mtcars$mpg ~ mtcars$am,
        main="Transmission Type Influence on Fuel Economy",
        xlab="Transmission Type (0 = automatic, 1 = manual)",
        ylab="Fuel Economy (miles per US gallon)")
```

#### Relative Feature Influence

The next figure presents the relative influence of each feature for the best fitted linear model. The linearity between the features is illustrated using a linear regression line.

```{r, results='hide'}
pairs(mtcars[,c('mpg', 'wt','qsec','am')], panel = function(x,y){
	points(x,y)
	abline(lm(y ~ x), col='red')
}, main="Motor Trend Car Best Features Scatterplot Mattrix")
```

#### Residuals

Finally, the following figure presents the residuals for the best fitted linear model. The Residuals vs Fitted shows no pattern, and the residuals appear linear on the Normal Q-Q plot emphasizing the validity of our model.

```{r, results='hide'}
par(mfrow = c(2, 2))
plot(fit_best)
```
